---
title: "Differential abundance of bacterial taxa at various taxonomic levels"
subtitle: "Using Maaslin3"
author: "Sarah Tanja"
date: 2026-03-16
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "" # Prevents appending '##' to beginning of lines in code output
)        
```

# Background

H~o~ : "PVC Leachate does not alter microbiome trajectory over
developmental time"

H~a~ : "PVC Leachate alters microbiome trajectory over developmental
time"

Microbiome Multivariable Associations with Linear Models (MaAsLin)

[MaAslin3](https://huttenhower.sph.harvard.edu/MaAsLin3) improves on
MaAslin2 by accounting for compositionality and accommodates
cross-sectional studies (that's ours!)

First read and get familiar with the: [MaAslin3 package
README](https://github.com/biobakery/maaslin3) & the [MaAslin3
tutorial](https://github.com/biobakery/biobakery/wiki/maaslin3) from the Stanford University Huttenhower Lab.

> William A. Nickols, Thomas Kuntz, Jiaxian Shen, Sagun Maharjan, Himel Mallick, Eric A. Franzosa, Kelsey N. Thompson, Jacob T. Nearing, Curtis Huttenhower. MaAsLin 3: Refining and extending generalized multivariable linear models for meta-omic association discovery. bioRxiv 2024.12.13.628459; doi: https://doi.org/10.1101/2024.12.13.628459

Here we aim to test changes in relative abundance (how many) and
prevalence (presence/absence) of specific taxa due to leachate, and the
interaction of leachate exposure over time, while controlling for spawn
night and read depth.

Our PERMANOVA detected that spawn night was a significant batch effect
and introduced nuisance variation into our data. To account for this
`(1 | spawn_night)` adds a random intercept for each unique spawn night
--- that is, each spawn night gets its own baseline shift in abundance,
but the effects of `hpf` and `leachate` is assumed to be the same across
nights.

[MaAsLin3 Wiki](https://github.com/biobakery/biobakery/wiki/maaslin3)
Any significant abundance associations with a categorical variable
should usually have at least 10 observations in each category.
Significant prevalence associations with categorical variables should
also have at least 10 samples in which the feature was present and at
least 10 samples in which it was absent for each significant category.
Significant abundance associations with continuous metadata should be
checked visually for influential outliers.

::: callout-important
There are also a few rules of thumb to keep in mind:

-   Models should ideally have about **10 times as many samples** (all
    samples for logistic fits, non-zero samples for linear fits) **as
    covariate terms** (**all continuous variables plus all categorical
    variable levels**).

```         
We have 63 samples... so the maximum number of terms we should use is 6 
```

-   Significant associations for MaAsLin 3 are results with no model
    fitting errors, and joint q-value less than 0.1

-   Significant abundance associations with continuous metadata should
    be checked visually for influential outliers.
:::

# Install MaAsLin3

```{r, eval=FALSE}
#library("devtools")
#install_github("biobakery/maaslin3")
```

# Load Libraries

```{r}
library(maaslin3)
library(qiime2R)
library(tidyverse)
```

> MaAsLin3 requires two input files, one for taxonomic or functional
> feature abundances, and one for sample metadata.

1.  Data (or features) file

-   This file is tab-delimited.
-   Formatted with features as columns and samples as rows.
-   The transpose of this format is also okay.
-   Possible features in this file include **microbes**, genes,
    pathways, etc.

2.  Metadata file

-   Formatted with metadata as columns and samples as rownames
-   Is a data.frame object
-   Includes per sample read counts

# Setup

## Set custom ggplot theme

```{r}
library(ggsidekick) # theme by sean anderson
theme_sleek_axe <- function() {
  theme_sleek() +
    theme(
      panel.border = element_rect(
        color = "grey70",
        fill = NA,
        linewidth = 0.5
      ),
      axis.line.x  = element_line(color = "grey70"),
      axis.line.y  = element_line(color = "grey70"),
      
      plot.title = element_text(
        size  = 10,
        color = "grey40",
        hjust = 0.5,
        margin = margin(b = 20)
      ),
      
      strip.text = element_text(size = 8, color = "grey40"),
  
      
      plot.subtitle = element_text(
        size   = 8,
        color  = "grey40",
        hjust  = 0.5,
        margin = margin(b = 40)
      ),
      
      axis.title = element_text(size = 8, color = "grey40"),
      axis.text = element_text(size = 7, color = "grey40"),
      legend.title = element_text(size = 8, color = "grey40"),
      legend.text = element_text(size = 7, color = "grey40")
    )
}

```

## Set colorschemes

```{r}
leachate.colors <- c(control = "#AEF1FF", 
                     low     = "#BBC7FF",
                     mid     = "#7D8BFF", 
                     high    = "#592F7D")

stage.5.colors <- c(egg           = "#FFE362",
                    cleavage      = "#EBA600", 
                    morula        = "#E6AA83",
                    prawnchip     = "#D9685B", 
                    earlygastrula = "#A2223C")

stage.3.colors <- c(cleavage      = "#EBA600", 
                    prawnchip     = "#D9685B", 
                    earlygastrula = "#A2223C")

status.colors <- c(typical   = "#75C165", 
                   uncertain = "#E3FAA5", 
                   malformed = "#8B0069")

night.colors <- c(July_6th = "#E3FAA5", 
                  July_7th = "#578B21", 
                  July_8th = "#1E2440")

night.colors.alt <- c(July_6th = "#21918C", 
                      July_7th = "#201158", 
                      July_8th = "#1E2440")
```

## Define file paths

```{r}
feature_table_path <- "../../salipante/241121_StonyCoral/270x200/"
metadata_path <- "../../metadata/meta.csv"
output_path <- "../../output/maaslin/"
fig_path <- "../../output/figs"
```

## Load metadata

```{r}
# Load metadata
metadata <- read_csv(metadata_path)

# set factors
metadata <- metadata %>% 
  mutate(
    collection_date = as.Date(collection_date, format = "%d-%b-%Y"),
    stage    = factor(stage,    levels = c("cleavage", "prawnchip", "earlygastrula"), ordered = TRUE),
    leachate = factor(leachate, levels = c("control", "low", "mid", "high"),        ordered = TRUE),
    spawn_night = factor(
      collection_date,
      levels  = as.Date(c("06-Jul-2024", "07-Jul-2024", "08-Jul-2024"), format = "%d-%b-%Y"),
      labels  = c("July 6th", "July 7th", "July 8th"),
      ordered = TRUE
    )
  )
```

## Put read depth per sample into the metadata

> Because MaAsLin 3 identifies prevalence (presence/absence)
> associations, sample read depth (number of reads) should be included
> as a covariate if available. Deeper sequencing will likely increase
> feature detection in a way that could spuriously correlate with
> metadata of interest when read depth is not included in the model.

```{r}
readepth <- read_csv("../../salipante/Sarah_StonyCoral/241121_StonyCoral_readcounts.csv")
```

```{r}
metadata <- metadata %>% 
  left_join(readepth, by = c("sample_id" = "sample"))
```

```{r}
# make samples rownames
meta <- metadata %>% 
  column_to_rownames(var = "sample_id") %>% 
  as.data.frame()

# View metadata structure
str(meta)
```

# Load Data

## Feature tables L4-7

```{r}
# L4 Order
## Load feature table from QIIME2 artifact
ft_l4 <- read_qza(file.path(feature_table_path, "collapsed-l4.qza"))$data

## Convert to data.frame
ft_l4 <- ft_l4 %>% 
  as.data.frame()

# L5 Family
## Load feature table from QIIME2 artifact
ft_l5 <- read_qza(file.path(feature_table_path, "collapsed-l5.qza"))$data

## Convert to data.frame
ft_l5 <- ft_l5 %>% 
  as.data.frame()

# L6 Genus
## Load feature table from QIIME2 artifact
ft_l6 <- read_qza(file.path(feature_table_path, "collapsed-l6.qza"))$data

## Convert to data.frame
ft_l6 <- ft_l6 %>% 
  as.data.frame()

# L7 Species
## Load feature table from QIIME2 artifact
ft_l7 <- read_qza(file.path(feature_table_path, "collapsed-l7.qza"))$data

## Convert to data.frame
ft_l7 <- ft_l7 %>% 
  as.data.frame()
```

# Run MaAslin3

- many taxa vary with time alone
- some taxa vary with leachate
- some taxa vary with leachate differently across time

**Model formula:**
`Abundance|Prevelance ~ leachate * hpf + reads + (1|spawn_night)`

**This model tests:**

-   Main effect of `leachate` (ordered categorical predictor; control, low, mid, high): Differences in the response between the levels of the `leachate` treatment, averaged over all `hpf`values. We choose to represent leachate categorically instead of a continuous variable here because we do not assume that response to increasing leachate levels is linear, and we want to capture any non-linear patterns of response across the different leachate treatments.

- Main effect of `hpf` (continuous): Overall change in the response across developmental time, averaged over all leachate levels. This tests if there is a general trend of response change with time regardless of leachate treatment.    

-   Interaction between `leachate` and `hpf`: Whether the slope of the response across hpf differs depending on the leachate category. This tests if `leachate` modifies how the response changes continuously over time; e.g., some leachate levels may accelerate or slow the effect of `hpf`.

-   Effect of `reads` (continuous covariate): Controls for the influence of read depth on the response, adjusting other effects accordingly.

-   Random intercept for `spawn_night`: Accounts for variability and non-independence among samples collected on the same spawn night by allowing different baseline response levels for each spawn night.

Interpretation-wise, if the interaction term is significant, it
indicates that the "trajectory" or slope of response change with `hpf`
differs by `leachate` group. If the interaction is not significant but
`leachate` is, then `leachate` affects overall response level but not
the shape of the response over time.

Graphically, imagine separate regression lines (response vs. `hpf`) for
each `leachate` group. The interaction tests whether these lines have
different slopes. The main effects test differences in intercepts and
general trend.

This model lets you hone in on how `leachate` influences response
patterns over continuous developmental time (`hpf`), adjusting for
sequencing depth (`reads`) and including random intercepts for
`spawn_night`.

- ordered_effects: A factored categorical variable to be included. Consecutive levels will be tested for significance against each other with contrast tests, and the resulting associations will correspond to effect sizes, standard errors, and significances of each level versus the previous.
Contrast tests are performed with multcomp::glht (fixed effects and logistic mixed effects) and lmerTest::contest (linear mixed effects).

## Ordered
```{r eval=FALSE, warning=FALSE, message=FALSE}
set.seed(03172026)
maaslin_fit <- maaslin3(
  input_data = ft_l7,
  input_metadata = meta,
  output = file.path(output_path, "ordered"),
  formula = ~ leachate*hpf + reads + (1|spawn_night),
  verbosity = 'ERROR',
  cores = 3, 
  plot_associations = FALSE, 
  save_models = TRUE
)
```

```{r, message=FALSE, warning=FALSE}
L7_sig <- read_tsv(file.path(output_path, "Interaction/L7_leachatehpf_cat.con/significant_results.tsv"))

L7_normalized<- read_tsv(file.path(output_path, "Interaction/L7_leachatehpf_cat.con/features/data_norm.tsv"))

L7_transformed <- read_tsv(file.path(output_path, "Interaction/L7_leachatehpf_cat.con/features/data_transformed.tsv"))

L7_filtered <- read_tsv(file.path(output_path, "Interaction/L7_leachatehpf_cat.con/features/filtered_data.tsv"))
```

```{r}
L7_sig <- L7_sig %>% 
  mutate(
    CI_lwr = coef - 1.98 * stderr,
    CI_upr = coef + 1.98 * stderr
  ) %>% 
  filter(is.na(error)) %>% 
  filter(metadata != "reads")

L7_feat <- unique(L7_sig$feature)
```

There are 57 unique features

```{r}
L7_sig %>% 
  group_by(model) %>%
  summarise(n = n_distinct(feature))
```

```{r}
# Extract only unique features to a character vector for prevalence and abundance models 
feat_prev <- L7_sig %>% 
  filter(model == "prevalence") %>% 
  distinct(feature) %>% 
  pull(feature)

feat_abun <- L7_sig %>% 
  filter(model == "abundance") %>% 
  distinct(feature) %>% 
  pull(feature)

# Which overlap?
feat_overlap <- intersect(feat_prev, feat_abun)

# How many unique features are in each model?
# Only in prevalence model
paste0("There are ", length(feat_prev), " total and ", length(setdiff(feat_prev, feat_abun)), " unique significant features in prevalence model")
# Only in abundance model
paste0("There are ", length(feat_abun), " total and ", length(setdiff(feat_abun, feat_prev)), " unique significant features in abundance model" )
# In both models
paste0("There are ", length(feat_overlap), " significant features in both")
```
::: callout-important
ALL taxa that are significant in the prevalence model are also significant in the abundance model, but there are some taxa that are only significant in the abundance model and not the prevalence model. This suggests that some taxa show significant changes in relative abundance across leachate and time, but do not show significant changes in presence/absence. In other words, these taxa may be consistently present across samples but vary in how abundant they are, rather than being present in some samples and absent in others. Meanwhile, all taxa that show significant changes in presence/absence also show significant changes in relative abundance, which makes sense because if a taxon is present in some samples and absent in others, it is likely to also show differences in how abundant it is across those samples.
:::


> In abundance models, a one-unit change in the metadatum variable
> corresponds to a 2coef fold change in the relative abundance of the
> feature.

> In prevalence models, a one-unit change in the metadatum variable
> corresponds to a coef change in the log-odds of a feature being
> present.

# Pivot data longer for plotting

##### Make function to extract taxa levels
```{r}
extract_tax_level <- function(x, rank) {
  # rank like "d", "p", "c", "o", "f", "g", "s"
  str_extract(x, paste0("(?<=", rank, "__)[^;]+"))
}
```

example to use funtion like: df2 \<- df %\>% mutate( domain =
extract_tax_level(feature, "d"), phylum = extract_tax_level(feature,
"p"), class = extract_tax_level(feature, "c"), order =
extract_tax_level(feature, "o"), family = extract_tax_level(feature,
"f"), genus = extract_tax_level(feature, "g"), species =
extract_tax_level(feature, "s") )

```{r}
L7_sig_long <- L7_normalized %>% 
  as_tibble() %>% 
  rename(sample_id = feature) %>% 
  pivot_longer(
    cols = 2:last_col(),
    names_to = "feature",
    values_to = "norm_abundance"
  ) %>% 
  filter(feature %in% L7_feat) %>%
  replace_na(list(norm_abundance = 0)) %>%
  left_join(metadata, by = "sample_id") %>% 
  mutate(
    order = extract_tax_level(feature, "o"),
    family = extract_tax_level(feature, "f"), 
    genus = extract_tax_level(feature, "g"),
    species = extract_tax_level(feature, "s"), 
    genus_species = paste0(genus, " ", species)
  ) %>% 
  mutate(presence = ifelse(norm_abundance > 0, 1, 0),
    prevalence = if_else(feature %in% feat_prev, "yes", "no"),
    abundance  = if_else(feature %in% feat_abun, "yes", "no")
  )

L7_feat_all <- unique(L7_sig_long$genus_species)
```

# Abundance

## Species level (L7)
```{r, fig.width = 35, fig.height = 30}
## Relative abundance (normalized read counts)
L7_sig_long %>% 
ggplot(., aes(x = factor(hpf), y = norm_abundance, color = leachate, fill = leachate)) +
  geom_boxplot(width = 0.6, alpha = 0.6, outlier.shape = NA, position = position_dodge(width = 0.8)) +
  geom_jitter(alpha = 0.8, position = position_dodge(width = 0.8)) +
  facet_wrap(~genus_species, scales = "free_y") +
  scale_fill_manual(values = leachate.colors) +
  scale_color_manual(values = leachate.colors) +
  labs(
      title = "Relative Abundance of leachate responsive bacterial Species in 14 hours of embryonic development",
      x = "Hours Post Fertilization",
      y = "Relative Abundance (normalized read counts)"
    ) +
  theme_sleek_axe()
```
```{r}
L7_sig_long %>%
  group_split(genus_species) %>%
  walk(function(df) {

    p <- ggplot(df, aes(x = factor(hpf), y = norm_abundance,
                        color = leachate, fill = leachate)) +
      geom_boxplot(width = 0.6, alpha = 0.6, outlier.shape = NA,
                   position = position_dodge(width = 0.8)) +
      geom_jitter(alpha = 0.8,
                  position = position_dodge(width = 0.8)) +
      scale_fill_manual(values = leachate.colors) +
      scale_color_manual(values = leachate.colors) +
      labs(
        title = paste("Relative abundance of", unique(df$genus_species)),
        x = "Hours Post Fertilization",
        y = "Relative Abundance (normalized read counts)"
      ) +
      theme_sleek_axe()

    ggsave(
      filename = file.path(fig_path, paste0("species_abundance_boxplots/L7_", unique(df$genus_species), ".png")),
      plot = p,
      width = 6,
      height = 4,
      dpi = 900
    )
    
    print(p)

  })
```

## Family (L5)
```{r}
L7_sig_long %>%
  group_split(family) %>%
  walk(function(df) {

    p <- ggplot(df, aes(x = factor(hpf), y = norm_abundance,
                        color = leachate, fill = leachate)) +
      geom_boxplot(width = 0.6, alpha = 0.6, outlier.shape = NA,
                   position = position_dodge(width = 0.8)) +
      geom_jitter(alpha = 0.8,
                  position = position_dodge(width = 0.8)) +
      scale_fill_manual(values = leachate.colors) +
      scale_color_manual(values = leachate.colors) +
      labs(
        title = paste("Relative abundance of", unique(df$family)),
        x = "Hours Post Fertilization",
        y = "Relative Abundance (normalized read counts)"
      ) +
      theme_sleek_axe()

    ggsave(
      filename = file.path(fig_path, paste0("family_abundance_boxplots/L7_", unique(df$family), ".png")),
      plot = p,
      width = 6,
      height = 4,
      dpi = 900
    )
    
    print(p)

  })
```

```{r, fig.width = 12, fig.height = 12}
## Relative abundance (normalized read counts)
L7_sig_long %>% 
ggplot(., aes(x = factor(hpf), y = norm_abundance, color = leachate, fill = leachate)) +
  geom_boxplot(width = 0.6, alpha = 0.6, outlier.shape = NA, position = position_dodge(width = 0.8)) +
  geom_jitter(alpha = 0.8, position = position_dodge(width = 0.8)) +
  facet_wrap(~family, scales = "free_y") +
  scale_fill_manual(values = leachate.colors) +
  scale_color_manual(values = leachate.colors) +
  labs(
      title = "Relative Abundance of leachate responsive bacterial Famiy in 14 hours of embryonic development",
      x = "Hours Post Fertilization",
      y = "Relative Abundance (normalized read counts)"
    ) +
  theme_sleek_axe()
```
```{r}
ggsave(
  filename = file.path(fig_path, "family_abundance.png"),
  width = 12,
  height = 12,
  dpi = 900
)
```

## Order (L4)
```{r}
L7_sig_long %>%
  group_split(order) %>%
  walk(function(df) {

    p <- ggplot(df, aes(x = factor(hpf), y = norm_abundance,
                        color = leachate, fill = leachate)) +
      geom_boxplot(width = 0.6, alpha = 0.6, outlier.shape = NA,
                   position = position_dodge(width = 0.8)) +
      geom_jitter(alpha = 0.8,
                  position = position_dodge(width = 0.8)) +
      scale_fill_manual(values = leachate.colors) +
      scale_color_manual(values = leachate.colors) +
      labs(
        title = paste("Relative abundance of", unique(df$order)),
        y = "Relative Abundance (normalized read counts)",
      ) +
      theme_sleek_axe()

    ggsave(
      filename = file.path(fig_path, paste0("order_abundance_boxplots/L7_", unique(df$order), ".png")),
      plot = p,
      width = 6,
      height = 4,
      dpi = 900
    )
    
    print(p)

  })
```

```{r, fig.width = 12, fig.height = 12}
## Relative abundance (normalized read counts)
L7_sig_long %>% 
ggplot(., aes(x = factor(hpf), y = norm_abundance, color = leachate, fill = leachate)) +
  geom_boxplot(width = 0.6, alpha = 0.6, outlier.shape = NA, position = position_dodge(width = 0.8)) +
  geom_jitter(alpha = 0.8, position = position_dodge(width = 0.8)) +
  facet_wrap(~order, scales = "free_y") +
  scale_fill_manual(values = leachate.colors) +
  scale_color_manual(values = leachate.colors) +
  labs(
      title = "Relative Abundance of leachate responsive bacterial Order in 14 hours of embryonic development",
      x = "Hours Post Fertilization",
      y = "Relative Abundance (normalized read counts)"
    ) +
  theme_sleek_axe()
```
```{r}
ggsave(
  filename = file.path(fig_path, "order_abundance.png"),
  width = 12,
  height = 12,
  dpi = 900
)
```

# Prevalence

## Species (L7)
```{r, fig.width = 14, fig.height = 30}
L7_sig_long %>% filter(feature %in% feat_prev) %>%
ggplot(.,
       aes(x = leachate,
           y = as.numeric(presence),
           color = leachate,
           fill = leachate)) +
    
  geom_smooth(
    aes(group = 1),
    method = "glm",
    method.args = list(family = "binomial"),
    se = FALSE,
    color = "grey40",
    fill = "grey70",
    linewidth = 0.8
  ) +
  geom_rug(
    aes(x = leachate_mgL),
    sides = "bt",
    color = "grey70",
    alpha = 0.5
  ) +
  geom_dotplot(
    binaxis = "y",
    stackdir = "centerwhole",
    alpha = 0.5,
    shape = 1,
    dotsize = 1.5
  ) +
    # mean points
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 4
  ) +

  facet_grid(genus_species~hpf) +
  coord_fixed(ratio = 2.8) +
  scale_fill_manual(values = leachate.colors) +
  scale_color_manual(values = leachate.colors) +
  labs(
    title = "Prevalence of leachate responsive bacteria Species across leachate levels and developmental stages",
    x = "Leachate Level",
    y = "Prevalence (proportion of samples with bacteria Species present)"
  ) +
  theme_sleek_axe()+ 
  theme(
  strip.text.y = element_text(size = 8, color = "grey40", angle = 0, hjust = 0)
)
```

```{r}
ggsave(
  filename = file.path(fig_path, "species_prevalence.png"),
  width = 12,
  height = 12,
  dpi = 900
)
```

## Family (L5)
```{r, fig.width = 14, fig.height = 30}
L7_sig_long %>% filter(feature %in% feat_prev) %>%
ggplot(.,
       aes(x = leachate,
           y = as.numeric(presence),
           color = leachate,
           fill = leachate)) +
    
  geom_smooth(
    aes(group = 1),
    method = "glm",
    method.args = list(family = "binomial"),
    se = FALSE,
    color = "grey40",
    fill = "grey70",
    linewidth = 0.8
  ) +
  geom_rug(
    aes(x = leachate_mgL),
    sides = "bt",
    color = "grey70",
    alpha = 0.5
  ) +
  geom_dotplot(
    binaxis = "y",
    stackdir = "centerwhole",
    alpha = 0.5,
    shape = 1,
    dotsize = 1.5
  ) +
    # mean points
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 4
  ) +

  facet_grid(family~hpf) +
  coord_fixed(ratio = 2.8) +
  scale_fill_manual(values = leachate.colors) +
  scale_color_manual(values = leachate.colors) +
  labs(
    title = "Prevalence of leachate responsive bacteria Family across leachate levels and developmental stages",
    x = "Leachate Level",
    y = "Prevalence (proportion of samples with bacteria Species present)"
  ) +
  theme_sleek_axe()+ 
  theme(
  strip.text.y = element_text(size = 8, color = "grey40", angle = 0, hjust = 0)
)
```


```{r}
ggsave(
  filename = file.path(fig_path, "family_prevalence.png"),
  width = 14,
  height = 30,
  dpi = 900)

```

## Order (L4)
```{r, fig.width = 12, fig.height = 18}
L7_sig_long %>% filter(feature %in% feat_prev) %>%
ggplot(.,
       aes(x = leachate,
           y = as.numeric(presence),
           color = leachate,
           fill = leachate)) +
    
  geom_smooth(
    aes(group = 1),
    method = "glm",
    method.args = list(family = "binomial"),
    se = FALSE,
    color = "grey40",
    fill = "grey70",
    linewidth = 0.8
  ) +
  geom_rug(
    aes(x = leachate_mgL),
    sides = "bt",
    color = "grey70",
    alpha = 0.5
  ) +
  geom_dotplot(
    binaxis = "y",
    stackdir = "centerwhole",
    alpha = 0.5,
    shape = 1,
    dotsize = 1.5
  ) +
    # mean points
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 4
  ) +

  facet_grid(order~hpf) +
  coord_fixed(ratio = 2.8) +
  scale_fill_manual(values = leachate.colors) +
  scale_color_manual(values = leachate.colors) +
  labs(
    title = "Prevalence of leachate responsive bacteria Order across leachate levels and developmental stages",
    x = "Leachate Level",
    y = "Prevalence (proportion of samples with bacteria Species present)"
  ) +
  theme_sleek_axe()+ 
  theme(
  strip.text.y = element_text(size = 8, color = "grey40", angle = 0, hjust = 0)
)
```

```{r}
ggsave(
  filename = file.path(fig_path, "order_prevalence.png"),
  width = 12,
  height = 18,
  dpi = 900
)
```

# Plot fitted models

- pull one fitted model for one taxon
- make a prediction grid from your metadata
- predict fitted probabilities
- plot observed presence + fitted curve/lines
- loop over taxa and save plots

```{r}
names(maaslin_fit$fit_data_prevalence$fits)[1:10]
```

I only want to plot the statistically significant taxa (those which are differentially prevalent). This list is a character vector named `feat_prev`

Subset list of model fits to `feat_prev`
```{r}
prev_fit_list <- maaslin_fit$fit_data_prevalence$fits

prev_fit_list_sig <- fit_list[feat_prev]
```


# Plot fitted prevalence model 
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

plot_maaslin_prev_fit <- function(feature_name, fit_obj, data_long) {
  
  # subset observed data for this feature
  df_feat <- data_long %>%
    filter(feature == feature_name)
  
  # create prediction grid
  newdata <- expand_grid(
    leachate = levels(factor(df_feat$leachate, ordered = is.ordered(df_feat$leachate))),
    hpf = unique(df_feat$hpf)
  ) %>%
    mutate(
      reads = median(df_feat$reads, na.rm = TRUE),
      spawn_night = df_feat$spawn_night[1]
    )
  
  # predict on response scale
  newdata <- newdata %>%
    mutate(
      pred = predict(fit_obj, newdata = newdata, type = "response", re.form = NA)
    )
  
  # plot
  ggplot(df_feat, aes(x = leachate, y = presence)) +
    geom_jitter(width = 0.1, height = 0.05, alpha = 0.5) +
    geom_line(
      data = newdata,
      aes(x = leachate, y = pred, group = hpf),
      inherit.aes = FALSE
    ) +
    geom_point(
      data = newdata,
      aes(x = leachate, y = pred),
      inherit.aes = FALSE,
      size = 2
    ) +
    facet_wrap(~ hpf) +
    labs(
      title = feature_name,
      x = "Leachate",
      y = "Presence probability"
    ) +
    theme_minimal()
}
```

# Summary 
```{r}
write_csv(L7_sig_long, file.path(output_path, "L7_significant_features_long.csv"))
```


```{r}
sessionInfo()
```

