---
title: "Differential abundance of bacterial taxa at various taxonomic levels"
subtitle: "Using Maaslin3"
author: "Sarah Tanja"
date: 2026-03-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
bibliography: "../microbiome_bibtex.bib"
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "" # Prevents appending '##' to beginning of lines in code output
)        
```

# Background

H~o~ : "PVC Leachate does not alter microbiome trajectory over
developmental time" H~a~ : "PVC Leachate alters microbiome trajectory
over developmental time"

Microbiome Multivariable Associations with Linear Models (MaAsLin)

[MaAslin3](https://huttenhower.sph.harvard.edu/MaAsLin3) improves on
MaAslin2 by accounting for compositionality and accommodates
cross-sectional studies (that's ours!)

First read and get familiar with the: [MaAslin3 package
README](https://github.com/biobakery/maaslin3) [MaAslin3
tutorial](https://github.com/biobakery/biobakery/wiki/maaslin3)

Here we aim to test changes in abundance (how many) and prevalence
(presence/absence) of specific taxa due to leachate, and their
interaction with stage while controlling for spawn night and read depth.

Our PERMANOVA detected that spawn night was a significant batch effect
and introduced nuisance variation into our data. To account for this
`(1 | spawn_night)` adds a random intercept for each unique spawn night
--- that is, each spawn night gets its own baseline shift in abundance,
but the effects of `hpf`/`stage` and `leachate` is assumed to be the
same across nights.

[MaAsLin3 Wiki](https://github.com/biobakery/biobakery/wiki/maaslin3)
Any significant abundance associations with a categorical variable
should usually have at least 10 observations in each category.
Significant prevalence associations with categorical variables should
also have at least 10 samples in which the feature was present and at
least 10 samples in which it was absent for each significant category.
Significant abundance associations with continuous metadata should be
checked visually for influential outliers.

::: callout-important
There are also a few rules of thumb to keep in mind:

-   Models should ideally have about **10 times as many samples** (all
    samples for logistic fits, non-zero samples for linear fits) **as
    covariate terms** (**all continuous variables plus all categorical
    variable levels**).

```         
We have 63 samples... so the maximum number of terms we should use is 6 
```
:::

Significant associations for MaAsLin 3 (no model fitting errors, joint
q-value less than 0.1)

# Install MaAsLin3

```{r, eval=false}
#library("devtools")
#install_github("biobakery/maaslin3")
```

# Load Libraries

```{r}
library(maaslin3)
library(qiime2R)
library(tidyverse)
```

> MaAsLin3 requires two input files, one for taxonomic or functional
> feature abundances, and one for sample metadata.

1.  Data (or features) file

-   This file is tab-delimited.
-   Formatted with features as columns and samples as rows.
-   The transpose of this format is also okay.
-   Possible features in this file include microbes, genes, pathways,
    etc.

2.  Metadata file

-   Formatted with metadata as columns and samples as rownames
-   Is a data.frame object
-   Includes per sample read counts

# Setup

```{r}
# Define file paths
feature_table_path <- "../../salipante/241121_StonyCoral/270x200/"
metadata_path <- "../../metadata/meta.csv"
output_path <- "../../output/maaslin/"
```

## Load metadata

```{r}
# Load metadata
meta <- read_csv(metadata_path)

# set factors
meta <- meta %>% 
  mutate(
    collection_date = as.Date(collection_date, format = "%d-%b-%Y"),
    stage    = factor(stage,    levels = c("cleavage", "prawnchip", "earlygastrula"), ordered = TRUE),
    leachate = factor(leachate, levels = c("control", "low", "mid", "high"),        ordered = TRUE),
    spawn_night = factor(
      collection_date,
      levels  = as.Date(c("06-Jul-2024", "07-Jul-2024", "08-Jul-2024"), format = "%d-%b-%Y"),
      labels  = c("July 6th", "July 7th", "July 8th"),
      ordered = TRUE
    )
  )
```

## Can I put read depth per sample into the metadata?

> Because MaAsLin 3 identifies prevalence (presence/absence)
> associations, sample read depth (number of reads) should be included
> as a covariate if available. Deeper sequencing will likely increase
> feature detection in a way that could spuriously correlate with
> metadata of interest when read depth is not included in the model.

As a counterpoint.. this adds complexity to the model and takes away
power.

```{r}
readepth <- read_csv("../../salipante/Sarah_StonyCoral/241121_StonyCoral_readcounts.csv")
```

```{r}
meta <- meta %>% 
  left_join(readepth, by = c("sample_id" = "sample"))
```

```{r}
# make samples rownames
meta <- meta %>% 
  column_to_rownames(var = "sample_id") %>% 
  as.data.frame()

# View metadata structure
str(meta)
```

Summarize number of samples representative of each variable of interest

```{r}
# Summarize number of samples per 'leachate' category
meta %>%
  group_by(leachate) %>%
  summarize(count = n())
```

```{r}
# Summarize number of samples per 'stage' category
meta %>%
  group_by(stage) %>%
  summarize(count = n())
```

```{r}
# Summarize number of samples per 'spawn_night' category
meta %>%
  group_by(spawn_night) %>%
  summarize(count = n())
```

# Load Data

## Feature tables L4-7

```{r}
# L4 Order
## Load feature table from QIIME2 artifact
ft_l4 <- read_qza(file.path(feature_table_path, "collapsed-l4.qza"))$data

## Convert to data.frame
ft_l4 <- ft_l4 %>% 
  as.data.frame()

# L5 Family
## Load feature table from QIIME2 artifact
ft_l5 <- read_qza(file.path(feature_table_path, "collapsed-l5.qza"))$data

## Convert to data.frame
ft_l5 <- ft_l5 %>% 
  as.data.frame()

# L6 Genus
## Load feature table from QIIME2 artifact
ft_l6 <- read_qza(file.path(feature_table_path, "collapsed-l6.qza"))$data

## Convert to data.frame
ft_l6 <- ft_l6 %>% 
  as.data.frame()

# L7 Species
## Load feature table from QIIME2 artifact
ft_l7 <- read_qza(file.path(feature_table_path, "collapsed-l7.qza"))$data

## Convert to data.frame
ft_l7 <- ft_l7 %>% 
  as.data.frame()
```

**Model formula:** `Abundance ~ leachate + timepoint + (1|cross)`

This tests: - Does leachate affect bacterial abundance (averaged across
all timepoints)? - Does timepoint affect bacterial abundance (averaged
across all leachate levels)?

::: callout-note
I modified the metadata file such that:

-   `leachate_mgL` is the numerical value (previously leachate) of
    microplastic mass in milligrams per liter of seawater

-   `hpf` is the numerical value (previously timepoint)

-   `parents` is the categorical string (previously cross) indicating
    pooled mix of colony crosses
:::

# Run MaAslin3

# Model Interactive

Categorical leachate & continuous time (interactive) **Model formula:**
`Abundance|Prevelance ~ leachate * hpf + reads + (1|spawn_night)`

**This model tests:** - Main effect of leachate (categorical):
Differences in the response between the levels of the leachate
treatment, averaged over all hpf values.

-   Main effect of hpf (continuous): The overall linear trend in the
    response as hours post-fertilization (hpf) changes, averaged over
    all leachate groups.

-   Interaction between leachate and hpf: Whether the slope of the
    response across hpf differs depending on the leachate category. This
    tests if leachate modifies how the response changes continuously
    over time; e.g., some leachate levels may accelerate or slow the
    effect of hpf.

-   Effect of reads (continuous covariate): Controls for the influence
    of read depth on the response, adjusting other effects accordingly.

-   Random intercept for spawn_night: Accounts for variability and
    non-independence among samples collected on the same spawn night by
    allowing different baseline response levels for each spawn night.

Interpretation-wise, if the interaction term is significant, it
indicates that the "trajectory" or slope of response change with hpf
differs by leachate group. If the interaction is not significant but
leachate is, then leachate affects overall response level but not the
shape of the response over time.

Graphically, imagine separate regression lines (response vs. hpf) for
each leachate group. The interaction tests whether these lines have
different slopes. The main effects test differences in intercepts and
general trend.

This model lets you hone in on how leachate influences response patterns
over continuous developmental time (hpf), adjusting for sequencing depth
and blocking by spawn night.

## L4 Order

```{r eval=FALSE, warning=FALSE, message=FALSE}
set.seed(03022026)
fit_Level4_Model1v1 <- maaslin3(
  input_data = ft_l4,
  input_metadata = meta,
  output = file.path(output_path, "Interaction/L4_interaction_cat.con"),
  formula = ~ leachate * hpf + reads + (1|spawn_night),
  verbosity = 'ERROR',
  cores = 3, 
  save_models = TRUE
)
```

Read in the significant results

```{r}
sig_L4 <- read_tsv(file.path(output_path, "Interaction/L4_interaction_cat.con/significant_results.tsv"))
```

Filter for leachate

```{r}
sig_pvc_L4 <- sig_L4 %>% 
  filter(metadata == "leachate") %>% 
  mutate(formula = "~ leachate * hpf + reads + (1|spawn_night)")

length(unique(sig_pvc_L4$feature))
```

::: callout-note
3 significant Orders

-   Nitrosococcales, linear = relative abundance

    -   More in leachate treatments compared to control

    -   ammonia-oxidizing bacteria (AOB)

        -   aerobes, they need oxygen!
        -   Elevated nitrogen waste (ammonia NH~3~/NH~4~^+^) may
            increase presence of AOB

        <!-- -->

        -   Increased organic matter turnover?

        <!-- -->

        -   System-level stress?

-   Parvibaculales, logistic = presence

-   Caulobacterales, logistic = presence

    -   
:::

Why might AOB increase in treatments with PVC leachate?

If embryos are stressed:

-   Protein turnover increases

-   Catabolism increases

-   Ammonia excretion increases

-   Embryos excrete ammonia directly into surrounding water.

So if leachate causes:

-   Metabolic stress

-   Oxidative stress

-   Increased cell death

-   Increased protein breakdown

That alone can drive enrichment.

This could be strongest at stages with higher metabolic rates (e.g.,
gastrula vs cleavage).

## L5 Family

```{r eval=FALSE, warning=FALSE, message=FALSE}
set.seed(03022026)
fit_Level5_Model1v1 <- maaslin3(
  input_data = ft_l5,
  input_metadata = meta,
  output = file.path(output_path, "Interaction/L5_interaction_cat.con"),
  formula = ~ leachate * hpf + reads + (1|spawn_night),
  verbosity = 'ERROR',
  cores = 3
)
```

Read in the significant results

```{r}
sig_L5 <- read_tsv(file.path(output_path, "Interaction/L5_interaction_cat.con/significant_results.tsv"))
```

Filter for leachate

```{r}
sig_pvc_L5 <- sig_L5 %>% 
  filter(metadata == "leachate") %>% 
  mutate(formula = "~ leachate * hpf + reads + (1|spawn_night)")

length(unique(sig_pvc_L5$feature))
unique(sig_pvc_L5$feature)
```

::: callout-note
4 Families

-   Rhizobiaceae, logistic

-   Burkholderiaceae, logistic

-   Parvibaculales PS1 clade, logistic

-   Nitrosococcales Methylophagaceae, linear

    -   NOT ammonia-oxidizing...

    -   grow on **small one-carbon or methylated compounds**

    -   Methylated dissolved organic compounds

    -   Leachate is a source for a mixture of dissolved organic carbon?

    -   Methylophagaceae specialize in metabolizing **C1 compounds** and
        methylated compounds that show up in marine DOM processing.
:::

## L6 Genus

```{r eval=FALSE, warning=FALSE, message=FALSE}
set.seed(03022026)
fit_Level6_Model1v1 <- maaslin3(
  input_data = ft_l6,
  input_metadata = meta,
  output = file.path(output_path, "Interaction/L6_interaction_cat.con"),
  formula = ~ leachate * hpf + reads + (1|spawn_night),
  verbosity = 'ERROR',
  cores = 3
)
```
