---
title: "Modeling predictions of bacterial taxa prevalence across leachate and time"
subtitle: "Can leachate predict the presence of taxa using GAMs?"
author: "Sarah Tanja"
date: 2026-03-17
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

# Background

Trying to model the prevalence of features across leachate and time is a
bit more complex than modeling abundance, but it can be done using the
same general approach. You can use generalized additive models (GAMs) to
model the probability of detection of a feature as a function of
leachate and time, while accounting for the non-linear relationships
between these variables and the response variable.

This was a suggestion by Amy Van Cise, she did something similar here:

> The repo you're looking for is:
> <https://github.com/MMARINeDNA/zDistribution>.
>
> You can find the scripts in the Scripts folder, and specifically you
> want to look at:
>
> -   02_Model_with_GAMs_Q1.R (model m1.2)
>
> <!-- -->
>
> -   02.1_Plot_H1_models.R (plots for model m1.2)
>
> <!-- -->
>
> -   02.2_m1.2_performance.R
>
>     Note that these aren't microbiome specific models or plots. Since
>     your data are binomial you can use the same model structure as
>     this one to model your relationship and predict binomial
>     probability outcomes. Or, if you'd simply like to add rug plots to
>     your ggplot object, you can see code for that in the plotting
>     script. -- Amy Van Cise

MaAsLin3 already ran a model , we just have to use it to make
predictions and then visualize the probability of detection on the Y
axis for our differentially prevalent taxa (features).

## What model is MaAsLin3 using for prevalence?

For **prevalence (presence/absence)**, MaAsLin3 fits a **generalized
linear mixed model (GLMM)**:

-    **Family:** binomial

-    **Link:** logit

-    **Form:** logistic regression (with optional random effects)

# Questions

#### Which taxa show leachate-associated changes beyond expected developmental change through time?

#### Does the probability of detection (POD) of a feature vary with leachate pollution level and across exposure/developmental time?

H~o~: Probability of detection does not vary across leachate or time

H~a~1: Probability of detection varies across leachate and time by ASV
feature

# Load packages

```{r}
library(tidyverse)
library(mgcv)
```

# Set paths

```{r}
input_path <- "../output/maaslin"
```

# Load data

```{r}
L7_sig_feat_long <- read_csv(file.path(input_path, "L7_significant_features_long.csv"))
```
