---
title: "NCBI SRA Submission"
author: "Sarah Tanja"
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "", # Prevents appending '##' to beginning of lines in code output
  dev = "jpeg",
  dpi = 600,
  fig.retina = 2)        
```

# Background

See [Submitting NGS sequences to NCBI SRA](https://sarahtanja.github.io/quarto-blog/posts/ncbi-sra-submission/ncbi-sra-submission.html) lab notebook post on instructions

each sample has 4 files to upload:

-   R1 fastq.gz

-   R1 fastq.gz.md5

-   R2 fastq.gz

-   R2 fastq.gz.md5

submission upload says it is missing files for samples `7C14` and `89C14`

```{r, engine='bash'}

ls ../data/*7C14*
```

```{r, engine='bash'}

ls ../data/*89C14*
```

The files are in our data folder so they must have not made the upload on the NCBI end

How many total files need to transfer?

```{r, engine='bash'}

ls -1 ../data/*fastq* | wc -l
```

252 files transferred! 4 files for each sample, and we have 63 samples.

::: callout-warning

**Warning:**

:   456L9_S42_R2_001.fastq.gz.md5 may be corrupted. If it is an archive, please re-compress it and upload it again.\
    Please double check that you had uploaded the correct file(s)
:::

```{r, engine='bash'}
cd ../data/
md5sum -c 456L9_S42_R2_001.fastq.gz.md5 #check if the md5sum is OK
md5sum 456L9_S42_R2_001.fastq.gz.md5 #write out checksum 
```

The checksum looks good, I just clicked 'submit' again and it went through.

Status is submitted awaiting processing. This SRA will be released on 2026-01-01 or upon publication.
