---
title: "Import, Trim, and Denoise ASVs"
author: "Sarah Tanja"
format: gfm
toc: true
toc-title: Contents <i class="bi bi-bookmark-heart"></i>
toc-depth: 5
toc-location: left
bibliography: "../microbiome_bibtex.bib"
reference-location: margin
citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

I'm re-running the commands already run by the Salipante lab (found in `../salipante/Sarah_StonyCoral/250405_QIIME2_filtered_StonyCloral_final.txt`)

# Load packages

```{r}
library(knitr)
library(dplyr)
library(stringr)
library(readr)
library(reticulate)
```

# Activate qiime2 environment in R

This is equivalent to `conda activate qiime2-2023.5` in the terminal but allows running qiime commands in r chunks with engine ='bash' throughout the quarto markdown doc.

```{r load-qiime-env, eval=TRUE}
use_condaenv(condaenv = "qiime2-2023.5", conda = "/home/shared/8TB_HDD_02/stanja/miniconda3/condabin/conda")

# Check successful env loading
py_config()
```

# Sequence files

-   Here I am working with raw `.fastq.gz` files located in the `../data/` directory
-   There are two `fastq.gz` files for each of the 63 samples, 1 forward read and 1 reverse read (R1 & R2)

# Metadata

-   metadata for the samples are located at `../metadata/metadata.tsv`

```{r}
metadata <- read_tsv("../metadata/metadata.tsv")
```

```{r, engine='bash'}
qiime metadata tabulate \
  --m-input-file ../metadata/metadata.tsv \
  --o-visualization ../output/metadata.qzv
```

# Modify read manifest

-   the read_manifest file the salipante lab used is here:

```{r}
read_manifest <- read_tsv("../input/241121_StonyCoral/250414_StonyCoral_read_manifest.tsv")

head(read_manifest)
```

The read_manifest is a list of the sample_id's and their corresponding forward and reverse fastq.gz file paths. In this directory, the fastq.gz files are located at `../data/`

However the paths must be absolute in order to run `qiime tools import` in the next step. Let's modify the read_manifest so that it works in this directory structure.

```{r}
normalizePath("../data/")
```

```{r, eval=TRUE}
read_manifest_mod <- read_manifest %>% 
  mutate(`forward-absolute-filepath` = str_replace(`forward-absolute-filepath`,
                                                  "^/mnt/labs/salipante/data/sequencing_runs/241121_StonyCoral/",
                                                  "/media/4TB_JPG_ext/stanja/gitprojects/coral-embryo-microbiome/data/"),
         `reverse-absolute-filepath` = str_replace(`reverse-absolute-filepath`,
                                                   "^/mnt/labs/salipante/data/sequencing_runs/241121_StonyCoral/",
                                                   "/media/4TB_JPG_ext/stanja/gitprojects/coral-embryo-microbiome/data/")
  )

head(read_manifest_mod)
```

```{r}
write_tsv(read_manifest_mod, "../input/read_manifest.tsv")
```

# Import data to QIIME

> `qiime tools import --type 'SampleData[PairedEndSequencesWithQuality]' --input-path 250414_StonyCoral_read_manifest.tsv --output-path 250414_StonyCoral_imported_reads.qza --input-format PairedEndFastqManifestPhred33V2 Imported 250414_StonyCoral_read_manifest.tsv as PairedEndFastqManifestPhred33V2 to 250414_StonyCoral_imported_reads.qza`

```{r, engine='bash'}
qiime tools import \
  --type 'SampleData[PairedEndSequencesWithQuality]' \
  --input-path ../input/read_manifest.tsv \
  --output-path ../output/imported_reads.qza \
  --input-format PairedEndFastqManifestPhred33V2 
```

# Trim

> `#klind primers so: qiime cutadapt trim-paired --i-demultiplexed-sequences 250414_StonyCoral_imported_reads.qza --p-cores 20 --p-front-f CCTACGGGNGGCWGCAG --p-front-r GACTACHVGGGTATCTAATCC --verbose --p-discard-untrimmed   --o-trimmed-sequences 250414_StonyCoral_imported_reads_trimmed.qza`

```{r, engine='bash'}
qiime cutadapt trim-paired \
  --i-demultiplexed-sequences ../output/imported_reads.qza \
  --p-cores 20 \
  --p-front-f CCTACGGGNGGCWGCAG \
  --p-front-r GACTACHVGGGTATCTAATCC \
  --verbose \
  --p-discard-untrimmed \
  --o-trimmed-sequences ../output/imported_reads_trimmed.qza
```

> `qiime demux summarize --i-data 250414_StonyCoral_imported_reads_trimmed.qza --o-visualization 250414_StonyCoral_imported_reads_trimmed.qzv`

```{r, engine='bash'}
qiime demux summarize \
  --i-data ../output/imported_reads_trimmed.qza \
  --o-visualization ../output/imported_reads_trimmed.qzv \

```

# Check quality scores

> `#qualities seemed fine, running across a number of trims to see which is best (qiime2-amplicon-2023.9) tron:/mnt/labs/salipante/data/projects/241121_StonyCoral/270x200 $ qiime dada2 denoise-paired --i-demultiplexed-seqs ../250414_StonyCoral_imported_reads_trimmed.qza --p-n-threads 20 --p-trunc-len-f 270 --p-trunc-len-r 200 --verbose --o-table 250414_StonyCoral_270x200_featureTable.qza --o-representative-sequences 250414_StonyCoral_240x230_representative-sequences.qza --o-denoising-stats 250414_StonyCoral_240x230_denoising-stats.qza`
