---
title: "Conda Environment Setup for QIIME2"
author: "Sarah Tanja"
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "", # Prevents appending '##' to beginning of lines in code output
  dev = "jpeg",
  dpi = 600,
  fig.retina = 2)        
```

# Install miniconda

Installation instruction for miniconda are [here](https://www.anaconda.com/docs/getting-started/miniconda/install#linux) - create a new directory named "miniconda3" in your home directory. - download the Linux Miniconda installation script for your chosen chip architecture and save the script as miniconda.sh in the miniconda3 directory. - run the miniconda.sh installation script in silent mode using bash. - remove the miniconda.sh installation script file after installation is complete.

```{r, engine='bash'}
echo "SHELL: $SHELL"
echo "whoami: $(whoami)"
echo "uname -a"
echo "pwd"
echo $0
ps -p $$    # what process is your current shell?
```

```{r}
cat /proc/version
```

> For Linux

```{r, engine='bash'}
mkdir -p ~/miniconda3
wget https://repo.anaconda.com/miniconda/Miniconda3-latest-Linux-x86_64.sh -O ~/miniconda3/miniconda.sh
bash ~/miniconda3/miniconda.sh -b -u -p ~/miniconda3
rm ~/miniconda3/miniconda.sh
```

> For Windows Powershell

```{r, engine='bash'}
Invoke-WebRequest -Uri "https://repo.anaconda.com/miniconda/Miniconda3-latest-Windows-x86_64.exe" -outfile ".\miniconda.exe"
Start-Process -FilePath ".\miniconda.exe" -ArgumentList "/S" -Wait
del .\miniconda.exe
```
> For Windows Command Prompt

```{r, engine='bash'}
curl https://repo.anaconda.com/miniconda/Miniconda3-latest-Windows-x86_64.exe -o .\miniconda.exe
start /wait "" .\miniconda.exe /S
del .\miniconda.exe
```

After installing Miniconda make sure you're running the latest version of `conda`.

```{r, engine='bash'}
conda update conda
```

# Install QIIME2 as a conda environment

Here I'm installing qiime2 version 2023.9 because it's the same one used by the Salipante Lab ...but it is using python 3.8 and picrust needs python 3.9 so there is a incompatible package conflict

## 2023.5

```{r create-qiime-2023.5, engine='bash'}
wget https://data.qiime2.org/distro/core/qiime2-2023.5-py38-linux-conda.yml
conda env create -n qiime2-2023.5 --file qiime2-2023.5-py38-linux-conda.yml
rm qiime2-2023.5-py38-linux-conda.yml
```

## 2023.9

::: callout-note
Use this version if possible because it's the same one the Salipante Lab is using!
:::

```{r create-qiime-2023.9, engine='bash'}
conda env create \
  --name qiime2-amplicon-2023.9 \
  --file https://raw.githubusercontent.com/qiime2/distributions/refs/heads/dev/2023.9/amplicon/released/qiime2-amplicon-ubuntu-latest-conda.yml
```

## 2024.4

```{r create-qiime-2024.4, engine='bash'}
conda env create \
  --name qiime2-amplicon-2024.2 \
  --file https://raw.githubusercontent.com/qiime2/distributions/refs/heads/dev/2024.2/amplicon/released/qiime2-amplicon-ubuntu-latest-conda.yml
```

# Activate qiime2 environment in R

```{r load-qiime-env, eval=TRUE}
use_condaenv(condaenv = "qiime2-amplicon-2023.9", conda = "/home/shared/8TB_HDD_02/stanja/miniconda3/condabin/conda")

# Check successful env loading
py_config()
```

If this is successful, the first line of output should show that the Python environment being used is the one in your conda environment path.

Your conda and qiime environments should be listed in your \$PATH:

```{r test-path, engine='bash'}
echo $PATH
```



# Troubleshooting





```{r, engine='bash'}
# Use the full path once to prove it exists
~/miniconda3/bin/conda --version

# Accept terms of service (tos)
# If needed, use full path: ~/miniconda3/bin/conda ...
~/miniconda3/bin/conda tos accept --override-channels --channel https://repo.anaconda.com/pkgs/main
~/miniconda3/bin/conda tos accept --override-channels --channel https://repo.anaconda.com/pkgs/r
```

> If you’re on Windows RStudio (Desktop) but Miniconda is on Linux/WSL... Your bash chunks might be using Windows Git-Bash, not the Linux bash that has ~/miniconda3. Point knitr to the Linux bash first.

```{r}
# Use WSL/Linux bash for bash chunks
knitr::opts_chunk$set(engine = "bash",
                      engine.path = list(bash = "/usr/bin/bash"))
```


```{r, engine='bash'}
echo "Bash path: $(command -v bash)"
uname -a
```

> Make conda available in all future {engine='bash'} chunks in this script (run this in an R chunk)

```{r}
# R code (not bash!)
writeLines('source "$HOME/miniconda3/etc/profile.d/conda.sh"', "~/.bash_conda_rc")
Sys.setenv(BASH_ENV = "~/.bash_conda_rc")
```

```{r, engine='bash'}
which bash
uname -a
echo "$HOME"
conda --version

```
```{r, engine='bash'}
set -e

echo "Bash: $(command -v bash)"
echo "HOME: $HOME"

# --- Sanity: do we have these files?
ls -l "$HOME/miniconda3" || true
ls -l "$HOME/miniconda3/bin/conda" "$HOME/miniconda3/condabin/conda" || true

# 1) Preferred: evaluate conda's shell hook into THIS shell
if [ -x "$HOME/miniconda3/bin/conda" ]; then
  echo "[1] Trying: conda shell hook"
  __conda_setup="$("$HOME/miniconda3/bin/conda" shell.bash hook 2>/dev/null)" && eval "$__conda_setup" && unset __conda_setup || true
fi

# 2) Fallback: source conda.sh if present
if ! command -v conda >/dev/null 2>&1 && [ -f "$HOME/miniconda3/etc/profile.d/conda.sh" ]; then
  echo "[2] Trying: source conda.sh"
  # shellcheck disable=SC1090
  source "$HOME/miniconda3/etc/profile.d/conda.sh"
fi

# 3) Last-resort: put conda *and* condabin on PATH
if ! command -v conda >/dev/null 2>&1; then
  echo "[3] Adding PATH fallbacks"
  export PATH="$HOME/miniconda3/condabin:$HOME/miniconda3/bin:$PATH"
fi

echo "conda on PATH? -> $(command -v conda || echo 'no')"
conda --version

```
```{r}
# Ensure knitr uses WSL bash explicitly
knitr::opts_chunk$set(engine.path = list(bash = "/usr/bin/bash"))

# Auto-init conda for every non-interactive bash spawned by R
# (bash reads BASH_ENV for non-interactive shells)
lines <- c(
  # 1) Try shell hook (best)
  '__conda_setup="$("$HOME/miniconda3/bin/conda" shell.bash hook 2>/dev/null)" && eval "$__conda_setup" && unset __conda_setup || true',
  # 2) Fallback to conda.sh
  '[ -f "$HOME/miniconda3/etc/profile.d/conda.sh" ] && source "$HOME/miniconda3/etc/profile.d/conda.sh" || true',
  # 3) PATH fallback
  'export PATH="$HOME/miniconda3/condabin:$HOME/miniconda3/bin:$PATH"'
)
writeLines(lines, "~/.bash_conda_rc")
Sys.setenv(BASH_ENV = "~/.bash_conda_rc")

```


```{r, engine='bash'}
# Tell conda to set itself up for future shells
~/miniconda3/bin/conda init bash

# Reload your shell config so `conda` works without restarting
source ~/.bashrc

# 3) Now this should work
conda update -n base conda

```

After installiong Miniconda make sure you're running the latest version of `conda`.

```{r, engine='bash'}
conda update conda
```
```{r, engine='bash'}
echo "which conda: $(command -v conda || echo 'NONE')"
ls -l "$HOME/miniconda3/bin/conda" "$HOME/miniconda3/condabin/conda" || true
head -n1 "$HOME/miniconda3/bin/conda" || true
ls -l "$HOME/miniconda3/bin/python"* || true

```

```{r}
# Ensure knitr uses WSL bash (you are, but explicit is fine)
knitr::opts_chunk$set(engine.path = list(bash = "/usr/bin/bash"))

# Put Miniconda first on PATH for all child processes (incl. bash chunks)
Sys.setenv(PATH = paste("/home/sarahtanja/miniconda3/bin",
                        "/home/sarahtanja/miniconda3/condabin",
                        Sys.getenv("PATH"), sep = ":"))

```