---
title: "Taxa Barplot Visualization from QIIME2"
author: "Sarah Tanja"
date: today
date-modified: today
format:
  gfm: default 
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

# Background

This document creates a publication-ready taxa barplot visualization from QIIME2 visualization files (.qzv). The script extracts taxonomic composition data from the QIIME2 artifacts and visualizes it using ggplot2.

## Inputs

- QIIME2 visualization file: `salipante/Sarah_StonyCoral/250414_StonyCoral_270x200_filtered_taxa-bar-plots.qzv`
- Metadata file: `metadata/metadata.tsv`

## Outputs

- Publication-ready taxa barplot figure
- Summary statistics of taxonomic composition

# Setup

## Load libraries

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(scales)
library(viridis)
library(RColorBrewer)
```

# Import Data

## Extract data from .qzv file

QIIME2 visualization files (.qzv) are zip archives containing CSV files with taxonomic data at different levels.

```{r}
# Define paths
qzv_file <- "../salipante/Sarah_StonyCoral/250414_StonyCoral_270x200_filtered_taxa-bar-plots.qzv"
temp_dir <- tempdir()
extract_dir <- file.path(temp_dir, "qzv_extract")

# Extract the .qzv file
unzip(qzv_file, exdir = extract_dir)

# Find the UUID directory (QIIME2 uses UUIDs for internal structure)
uuid_dir <- list.dirs(extract_dir, recursive = FALSE, full.names = TRUE)[1]
data_dir <- file.path(uuid_dir, "data")

# List available taxonomic levels
available_files <- list.files(data_dir, pattern = "\\.csv$")
cat("Available taxonomic levels:\n")
print(available_files)
```

## Load taxonomic data at different levels

```{r}
# Load level 7 (most detailed, species level)
level7_file <- file.path(data_dir, "level-7.csv")
taxa_l7 <- read_csv(level7_file, show_col_types = FALSE)

# Load level 6 (genus level) - often better for visualization
level6_file <- file.path(data_dir, "level-6.csv")
taxa_l6 <- read_csv(level6_file, show_col_types = FALSE)

# Load level 5 (family level)
level5_file <- file.path(data_dir, "level-5.csv")
taxa_l5 <- read_csv(level5_file, show_col_types = FALSE)

cat("Level 7 dimensions:", dim(taxa_l7), "\n")
cat("Level 6 dimensions:", dim(taxa_l6), "\n")
cat("Level 5 dimensions:", dim(taxa_l5), "\n")
```

## Load metadata

```{r}
metadata <- read_tsv("../metadata/metadata.tsv", 
                     skip = 1,  # Skip the q2:types line
                     show_col_types = FALSE)

# Display metadata structure
head(metadata)
cat("\nSample groups:\n")
table(metadata$leachate, metadata$timepoint)
```

# Data Processing

## Transform data for visualization

We'll use Level 6 (genus level) as it provides a good balance between detail and interpretability.

```{r}
# Transform data to long format
taxa_long <- taxa_l6 %>%
  pivot_longer(
    cols = -index,
    names_to = "taxon",
    values_to = "count"
  ) %>%
  rename(sample_id = index)

# Calculate relative abundance
taxa_long <- taxa_long %>%
  group_by(sample_id) %>%
  mutate(
    total_count = sum(count),
    rel_abundance = count / total_count * 100
  ) %>%
  ungroup()

# Extract taxonomic rank information
taxa_long <- taxa_long %>%
  mutate(
    # Extract genus (last named rank before __)
    genus = str_extract(taxon, "g__([^;_]+)", group = 1),
    # Extract family
    family = str_extract(taxon, "f__([^;_]+)", group = 1),
    # Extract phylum
    phylum = str_extract(taxon, "p__([^;_]+)", group = 1),
    # Create a simplified label
    genus = if_else(is.na(genus) | genus == "", 
                    paste0("Unknown_", family), 
                    genus)
  )

# Join with metadata
taxa_long <- taxa_long %>%
  left_join(metadata, by = c("sample_id" = "sample-id"))

head(taxa_long)
```

## Select top taxa for visualization

```{r}
# Calculate mean relative abundance per genus
top_genera <- taxa_long %>%
  group_by(genus) %>%
  summarize(mean_abundance = mean(rel_abundance)) %>%
  arrange(desc(mean_abundance)) %>%
  slice_head(n = 15) %>%
  pull(genus)

# Create a factor for plotting (top genera + "Other")
taxa_plot <- taxa_long %>%
  mutate(
    genus_grouped = if_else(genus %in% top_genera, genus, "Other")
  ) %>%
  group_by(sample_id, genus_grouped, development.stage, 
           timepoint, leachate) %>%
  summarize(rel_abundance = sum(rel_abundance), .groups = "drop")

# Order samples by metadata
taxa_plot <- taxa_plot %>%
  mutate(
    sample_id = factor(sample_id),
    genus_grouped = factor(genus_grouped, 
                          levels = c(top_genera, "Other"))
  )
```

# Visualization

## Taxa barplot - Main Figure

```{r}
#| fig-width: 12
#| fig-height: 8
#| fig-dpi: 300

# Create color palette
n_colors <- length(top_genera) + 1
colors <- c(brewer.pal(min(12, n_colors-1), "Set3"), "grey80")

# Create the barplot
p1 <- ggplot(taxa_plot, 
             aes(x = sample_id, y = rel_abundance, fill = genus_grouped)) +
  geom_bar(stat = "identity", width = 0.9) +
  scale_fill_manual(values = colors, name = "Genus") +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Bacterial Community Composition",
    subtitle = "Top 15 genera across developmental stages and leachate treatments",
    x = "Sample",
    y = "Relative Abundance (%)"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.5, "cm"),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11)
  )

print(p1)
```

## Taxa barplot - Faceted by Treatment

```{r}
#| fig-width: 14
#| fig-height: 10
#| fig-dpi: 300

# Aggregate by treatment groups
taxa_treatment <- taxa_long %>%
  mutate(
    genus_grouped = if_else(genus %in% top_genera, genus, "Other"),
    leachate_label = paste0("Leachate: ", leachate),
    timepoint_label = paste0(timepoint, " hpf")
  ) %>%
  group_by(leachate_label, timepoint_label, genus_grouped) %>%
  summarize(mean_rel_abundance = mean(rel_abundance), .groups = "drop") %>%
  mutate(
    genus_grouped = factor(genus_grouped, levels = c(top_genera, "Other"))
  )

# Create faceted barplot
p2 <- ggplot(taxa_treatment, 
             aes(x = timepoint_label, y = mean_rel_abundance, fill = genus_grouped)) +
  geom_bar(stat = "identity", width = 0.8) +
  scale_fill_manual(values = colors, name = "Genus") +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  facet_wrap(~ leachate_label, ncol = 4) +
  labs(
    title = "Bacterial Community Composition by Leachate Treatment and Timepoint",
    x = "Timepoint (hours post fertilization)",
    y = "Mean Relative Abundance (%)"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    legend.position = "bottom",
    legend.key.size = unit(0.4, "cm"),
    strip.background = element_rect(fill = "lightgray", color = "black"),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  guides(fill = guide_legend(nrow = 2))

print(p2)
```

## Taxa barplot - Phylum Level

```{r}
#| fig-width: 12
#| fig-height: 8
#| fig-dpi: 300

# Calculate top phyla
top_phyla <- taxa_long %>%
  group_by(phylum) %>%
  summarize(mean_abundance = mean(rel_abundance)) %>%
  arrange(desc(mean_abundance)) %>%
  slice_head(n = 10) %>%
  pull(phylum)

# Aggregate by phylum
taxa_phylum <- taxa_long %>%
  mutate(
    phylum_grouped = if_else(phylum %in% top_phyla, phylum, "Other")
  ) %>%
  group_by(sample_id, phylum_grouped, development.stage, 
           timepoint, leachate) %>%
  summarize(rel_abundance = sum(rel_abundance), .groups = "drop") %>%
  mutate(
    phylum_grouped = factor(phylum_grouped, levels = c(top_phyla, "Other"))
  )

# Create phylum-level barplot
p3 <- ggplot(taxa_phylum, 
             aes(x = sample_id, y = rel_abundance, fill = phylum_grouped)) +
  geom_bar(stat = "identity", width = 0.9) +
  scale_fill_brewer(palette = "Paired", name = "Phylum") +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Bacterial Community Composition at Phylum Level",
    x = "Sample",
    y = "Relative Abundance (%)"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )

print(p3)
```

# Summary Statistics

## Top taxa across all samples

```{r}
# Calculate overall abundance statistics
top_taxa_summary <- taxa_long %>%
  group_by(genus) %>%
  summarize(
    mean_abundance = mean(rel_abundance),
    sd_abundance = sd(rel_abundance),
    max_abundance = max(rel_abundance),
    n_samples_present = sum(rel_abundance > 0)
  ) %>%
  arrange(desc(mean_abundance)) %>%
  slice_head(n = 20)

print(top_taxa_summary)
```

## Diversity by treatment

```{r}
# Calculate diversity metrics per sample
diversity_metrics <- taxa_long %>%
  filter(count > 0) %>%
  group_by(sample_id, leachate, timepoint, development.stage) %>%
  summarize(
    richness = n(),  # Number of unique taxa
    shannon_entropy = -sum((rel_abundance/100) * log(rel_abundance/100)),
    .groups = "drop"
  )

# Summary by treatment
diversity_summary <- diversity_metrics %>%
  group_by(leachate, timepoint) %>%
  summarize(
    mean_richness = mean(richness),
    sd_richness = sd(richness),
    mean_shannon = mean(shannon_entropy),
    sd_shannon = sd(shannon_entropy),
    .groups = "drop"
  )

print(diversity_summary)
```

## Visualization of diversity metrics

```{r}
#| fig-width: 10
#| fig-height: 6

p4 <- ggplot(diversity_metrics, 
             aes(x = factor(timepoint), y = richness, 
                 color = factor(leachate), group = factor(leachate))) +
  geom_point(size = 3, alpha = 0.6) +
  geom_line() +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  scale_color_viridis_d(name = "Leachate\nConcentration") +
  labs(
    title = "Taxonomic Richness Across Treatments",
    x = "Timepoint (hours post fertilization)",
    y = "Number of Genera"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )

print(p4)
```

# Save Figures

```{r}
# Create output directory if it doesn't exist
dir.create("../output/figures", recursive = TRUE, showWarnings = FALSE)

# Save figures
ggsave("../output/figures/taxa_barplot_all_samples.png", 
       plot = p1, width = 12, height = 8, dpi = 300)

ggsave("../output/figures/taxa_barplot_by_treatment.png", 
       plot = p2, width = 14, height = 10, dpi = 300)

ggsave("../output/figures/taxa_barplot_phylum.png", 
       plot = p3, width = 12, height = 8, dpi = 300)

ggsave("../output/figures/diversity_by_treatment.png", 
       plot = p4, width = 10, height = 6, dpi = 300)

cat("Figures saved to output/figures/\n")
```

# Session Info

```{r}
sessionInfo()
```
