---
title: "PICRUSt2 pathway resulst"
author: "Sarah Tanja"
date: 2026-04-30
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = FALSE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output)        
```

# Background
> Predicted pathway abundances were inferred from ASV sequences using PICRUSt2 and compared among taxa associated with PVC leachate exposure
> You are modeling predicted function from taxonomy, so you are not directly measuring function but rather inferring it based on the taxonomy of the microbes present. This is a common approach in microbiome studies when direct functional measurements (like metagenomics or metatranscriptomics) are not available.
>“Predicted microbial functional potential shifted…”

note to self.. can't use qiime2R on windows so go to raven but can't update raven r version (needed for maaslin) so head back to minerva to run maaslin3 
# Load libraries

```{r, eval=TRUE}
library(reticulate)
library(tidyverse)
library(qiime2R)
library(maaslin3)
```


```{r}
path <- read_qza("../../output/picrust2/pathway_abundance.qza")$data
```

```{r}
ko <- read_qza("../../output/picrust2/ko_metagenome.qza")$data
```

## Define file paths

```{r}
metadata_path <- "../../metadata/meta.csv"
output_path <- "../../output/picrust2/"
fig_path <- "../../output/figs"
```

## Load metadata

```{r}
# Load metadata
metadata <- read_csv(metadata_path)

# set factors
metadata <- metadata %>% 
  mutate(
    collection_date = as.Date(collection_date, format = "%d-%b-%Y"),
    stage    = factor(stage,    levels = c("cleavage", "prawnchip", "earlygastrula"), ordered = TRUE),
    leachate = factor(leachate, levels = c("control", "low", "mid", "high"),        ordered = TRUE),
    spawn_night = factor(
      collection_date,
      levels  = as.Date(c("06-Jul-2024", "07-Jul-2024", "08-Jul-2024"), format = "%d-%b-%Y"),
      labels  = c("July 6th", "July 7th", "July 8th"),
      ordered = TRUE
    )
  )
```

```{r}
# make samples rownames
meta <- metadata %>% 
  column_to_rownames(var = "sample_id") %>% 
  as.data.frame()

# View metadata structure
str(meta)
```

# Pathways
## Transpose so rownames are samples and columns are pathways
```{r}
path <- t(path)
```

##  Run MaAsLin3
```{r}
set.seed(04302026)
mp <- maaslin3(
  input_data = path,
  input_metadata = meta,
  output = file.path(output_path, "maaslin_path"),
  formula = ~ leachate*hpf + (1|spawn_night),
  verbosity = 'ERROR',
  cores = 3, 
  plot_associations = FALSE, 
  save_models = TRUE
)
```

```{r}
sig_paths <- read_tsv(file.path(output_path, "maaslin_path/significant_results.tsv"))
```

.L Does the response rise steadily?
.Q Does it peak in middle doses?
.C Does it show a more complex non-monotonic pattern?

leachate 
```{r}
sig_leachate <- sig_paths %>% 
  filter(metadata == "leachate") %>% 
  arrange(qval_joint)

unique(sig_leachate$feature) 
```

Lipid metabolism pathways → membrane stress responses
Aromatic degradation pathways → breakdown of plastic-derived organics
Redox / respiration pathways → oxidative stress handling
 
- PWY-5179
Likely aromatic compound degradation
Breaks down organic compounds

# KOs

Transpose so rownames are samples and KOs are columns
```{r}
ko <- t(ko)
```

##  Run MaAsLin3
```{r}
set.seed(04302026)
ko <- maaslin3(
  input_data = ko,
  input_metadata = meta,
  output = file.path(output_path, "maaslin_ko"),
  formula = ~ leachate*hpf + (1|spawn_night),
  verbosity = 'ERROR',
  cores = 3, 
  plot_associations = FALSE, 
  save_models = TRUE
)
```

# ggpicrust2 package for visualizing results
```{r}
library(ggpicrust2)
```

# PCA of pathway matrix 
> Goal: Perform PCA analysis on pathway abundance data and create an informative visualization that includes a scatter plot of the first two principal components (PC1 vs PC2) with density plots for both PCs. The plot helps to visualize the clustering patterns and distribution of samples across different groups.

## Filter down to sig paths
```{r}

```



```{r}
# Perform PCA on pathway abundance data
pca_comp <- prcomp(path, scale. = TRUE)

# Create a data frame for plotting
pca_df <- as.data.frame(pca_comp$x)

# Add metadata to the PCA data frame
pca_df <- cbind(pca_df, meta)

# Set factors 
pca_df <- pca_df %>% 
  mutate(stage = factor(stage, levels = c("cleavage", "prawnchip", "earlygastrula")),
         leachate = factor(leachate, levels = c("control", "low", "mid", "high")))

# Calculate %Var in PC1 and PC2
percentVar <- (pca_comp$sdev^2) / sum(pca_comp$sdev^2) * 100

pc1_var <- round(percentVar[1], 1)
pc2_var <- round(percentVar[2], 1)
```

Labels for grouping variables
```{r}
labs_leachate <- c(control = "0 mg/L (control)",
                   low = "0.01 mg/L (low)",
                   mid = "0.1 mg/L (mid)",
                   high = "1 mg/L (high)")

labs_stage <- c(cleavage = "Cleavage", prawnchip = "Prawn chip", earlygastrula = "Early gastrula")
```


```{r}
# Create a scatter plot of PC1 vs PC2 with density plots
ggplot(pca_df, aes(x = PC1, y = PC2, color = leachate)) +
  geom_point(size = 3) +
  #geom_density_2d() +
  labs(title = "PCA of Pathway Abundance Data",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  theme(legend.title = element_blank())
```




